---
title: "Big Data in R with Arrow"
subtitle: "A microlearning module on larger-than-memory data analysis in R with Apache Arrow"
author: "[Pete Lawson, PhD]{.focus} | Data and Visualization Librarian"
---

Apache Arrow, and the `arrow` package in R, is an extremely fast toolbox for larger-than-memory data analysis.

### Reading and writing Apache Parquet files

Reading data in as parquet will offer a significant speedup. 

If you have data loaded from csv, you can write it into a parquet file.

To load the data from a single csv, you can use `open_csv_dataset("path/to/csv")`, which is a wrapper to `open_dataset()`.

```{r}
#| eval: FALSE
taxi_data <- open_csv_dataset("data/nyc-taxi/yellow_taxi_data.csv")
```

To take full advantage of parquet performance you can partition your file into individual parquet files, but you must be cautious:

Partition your data into files [between 20 MB and 2 GB]{.focus}, with no more than [10,000 partitions]{.focus}. Too small and reading metadata from each file will slow us down, to big and we won't see a performance advantage over one file.

Use "hive-partitioning" when saving to split our data into multiple directories based on key variable(s) of interest In this example we will save the NYC taxi data into parquet, partitioned on the variables `year` and `month`. 

Because the variables are saved in the directory names, Arrow will automatically infer the Parquet structure.

```{r}
#| eval: FALSE
nyc_taxi |>
  write_dataset(
    "nyc-taxi/",
    format = "parquet",
    partitioning = c("year", "month"),
    hive_style = TRUE
  )
```

We partitioned our single csv into [196]{.focus} parquet files, around [200 MB]{.focus} each, so we should see great performance!

### When to pull data into memory

Arrow uses `lazy evaluation`. Your `dplyr` verbs aren't executed until you explicitly tell Arrow to do so. 

You can tell it two ways: 

`collect()`: Arrow query into memory as a `tibble` (thousands of rows)
`compute()`: Arrow query into memory as an Arrow Table (millions of rows)

Use `nrow()` liberally. This will tell you how many rows will be returned [if]{.focus} you choose to bring the data into memory. Too many (more than a few million rows) and you likely need to filter the data down some more.

```{r}
#| eval: FALSE
nyc_taxi |>
  filter(year %in% 2020:2026) |>
  nrow()
```


### Schemas - here be dragons

When you read in a dataset in Arrow, it scans approximately 1 MB of the data for a csv, or the first file when reading parquet files, and guesses (imputes) the data types. If not all of your data is represented in that sample, Arrow may [infer the wrong type]{.focus}!

Be sure to check your schema using `schema(nyc_taxi)`. The `schema()` function can extract a schema from an object, but it can also be used to update a schema if yours is incorrect.

### Learn more 

View the `arrow` package [documentation](https://arrow.apache.org/docs/18.0/r/index.html), or work through this [excellent tutorial](https://posit-conf-2024.github.io/arrow/) from posit::conf(2025).


